---
title: "Working with 2022 microdata"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
urlcolor: blue
vignette: >
  %\VignetteIndexEntry{Working with 2022 microdata}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = identical(tolower(Sys.getenv("NOT_CRAN")), "true"),
  out.width = "100%"
)
```

## Why the 2022 census is different

For every census up to 2010, the microdata are openly available, and **{censobr}**
downloads them for you already processed. The sample microdata of the 2022
census are not: IBGE releases them as *dados controlados*, under controlled
access. **{censobr}** is therefore not allowed to redistribute them, and there is
no file for the package to download on your behalf.

What the package can do is read the file **you** obtained from IBGE and store it
in the local **{censobr}** cache, in the same format and under the same naming
convention as every other census year. That is what
`import_microdata22_controlado()` is for.

## Step 1. Get the data from IBGE

Request and download the 2022 sample microdata from IBGE at
[https://microdados.ibge.gov.br/](https://microdados.ibge.gov.br/). Download the
version of the data saved in **`.csv`** format.

The file you receive is a single `.zip` holding one subdirectory per state, each
with four tables:

| File in the zip | Content |
| --- | --- |
| `Domicilios` | households |
| `Familia` | families |
| `Mortalidade` | mortality |
| `Pessoas` | population |

## Step 2. Import it

Once you have downloaded the original zip file with the data, you only need to
pass the path to the zip file to `import_microdata22_controlado()`. Here's an 
example using a tiny *fake* data set with the same structure as the IBGE
file.

```{r, eval = FALSE}
library(censobr)

# path to the zip file with the raw data 
fake_zip <- system.file(
  "extdata/microdata_2022_controlado_fake.zip",
  package = "censobr"
  )

censobr::import_microdata22_controlado(zip_path = fake_zip)

```

The function unzips the file, read each of the four tables, adds
geography columns following the {geobr} convention, and save the
data sets as `.parquet` files in the **{censobr}** cache directory. 
Depending on your machine this takes a few minutes, most of it 
spent on the `Pessoas` table.


And the data can be read back with `{censobr}` using the familiar verbs:

```{r, eval = FALSE}
pop <- censobr::read_population(year = 2022)

fam <- censobr::read_families(year = 2022)

hou <- censobr::read_households(year = 2022)

mor <- censobr::read_mortality(year = 2022)

```


## Keep the original zip file

You only need to import the data **once**. After that the tables sit in the
cache and are read from disk, with no further processing.

There is one case where you have to import them again. The **{censobr}** cache is
versioned by data release, and each release lives in its own subdirectory. When a
new version of the package points to a newer data release, it will not find the
files you imported under the previous one.

So **store the original `.zip` from IBGE somewhere safe**. It is the only copy
you have, the package cannot download it for you, and you will need it if
you upgrade **{censobr}** or move to another machine. Deleting your cache with
`censobr_cache(delete_file = "all")` has the same consequence.

